Using HMMs for pre-snap coverage detection

### 2024-11-28 submission.Rmd
.

Choosing between man- and zone coverage is one of the most important strategic decisions a defensive coordinator has to take prior to each play in American football. While experienced coaches and quarterbacks can often identify these defensive strategies visually, the growing availability of tracking data presents another opportunity to infer the underlying tactic.
This project aims to leverage hidden Markov models (HMMs) to detect defensive strategies --- man or zone coverage --- based on pre-snap player movement data. 
By modeling hidden states that represent the offensive player to be guarded, this approach provides a data-driven framework for unraveling the complexity of defensive patterns and enables real-time tactical insights for coaching and analysis.

## Data

This video displays a touchdown from the Kansas City Chiefs against the Arizona Cardinals in Week 1 of the 2022 NFL season. We can see that, pre-snap, Mecole Hardman (KC #17) travels from the left side to the right side, followed by the defender Marco Wilson (AZ #20), i.e.\ a clear indication for man-coverage.

@@

@@

#### 1. Pre-motion analysis
We train a model to predict whether the defense plays a man- or zone coverage scheme. The 
model uses the previously described features, blablabla and as observations information from those plays without pre-snap motion and information until the first motion for plays with motion.

#### 2. HMM analysis
We model the movements of defensive players during the phase of pre-snap motion within a hidden Markov framework, in which the underlying states represent the offensive players to be guarded (see Franks et al., 2015 for a similar approach in basketball). In contrast to Groom et al., 2024, who enforce a state to proxy zone coverage during corner kicks in soccer, we cannot proceed similarly as the classical coverage zones in American football will only be covered post-snap. To remediate this, we derive the decision of man- or zone coverage from the number of switches for individual players. In particular, a low number of switches when offensive players are in motion indicates man coverage whereas a higher number indicates zone coverage.

@@

@@

>>>>

Choosing between man- and zone coverage is one of the most important strategic decisions a defensive coordinator has to take prior to each play in American football. While experienced coaches and quarterbacks can often identify these defensive strategies visually, the growing availability of tracking data presents another opportunity to infer the underlying tactic.
This project aims to leverage hidden Markov models (HMMs) to detect defensive strategies --- man or zone coverage --- based on pre-snap player movement data. 
By modeling hidden states that proxy the offensive player to be guarded, this approach provides a data-driven framework for unraveling the complexity of defensive patterns and enables real-time tactical insights for coaching and analysis.

The following video displays a touchdown from the Kansas City Chiefs against the Arizona Cardinals in Week 1 of the 2022 NFL season. We can see that, pre-snap, Mecole Hardman (KC #17) is in motion. He is immediately followed by the defender Marco Wilson (AZ #20), which is a clear indication for man-coverage.

@@

## Data
Analyzing tracking data from nine weeks of the NFL 2022 season, we aim to forecast the defensive scheme (man- or zone defense). For this, we use the corresponding data from PFF that analysed every play and assigned the categories \textit{man}, \textit{zone} and \textit{others} representing the different schemes. As it is not properly described what \textit{others} means, we omit every play that is associated with this value. Then, we end up with XY plays in total, from which the defense played Y in zone and X in man coverage.

Within these plays, we concentrate on the tracking data after the line has been set (because we are not interested in how players come out of the huddle) and before the ball has been snapped by the Center. For the HMM analysis, we further concentrate on those plays with pre-snap motion (ZZ plays).

##### Feature engineering
To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps:

\textbf{UMSCHREIBEN} \textit{For all players and the ball carrier we use the features already contained in the tracking data, namely x- and y-coordinates, speed, acceleration, distance covered, orientation and direction. Except for the ball carrier we further compute the euclidean distance, x-distance and y-distance to the ball carrier. For defensive players only, we additionally compute the absolute difference of the defender’s direction and the angle of the shortest segment between the defender and the ball carrier. Subsequently, we order all players (in each frame) with respect to their euclidean distance to the ball carrier and standardize all features.
More details regarding data preprocessing can be found in the Appendix.}

@@

#### 1. Pre-motion analysis
We train a model to predict whether the defense plays a man- or zone coverage scheme. The 
model uses the previously described features, blablabla.

#### 2. HMM analysis
We model the movements of defensive players during the phase of pre-snap motion within a hidden Markov framework, in which the underlying states represent the offensive players to be guarded (see Franks et al., 2015 for a similar approach in basketball). In contrast to Groom et al., 2024, who enforce a state to proxy zone coverage during corner kicks in soccer, we cannot proceed similarly as the classical coverage zones in American football will only be covered by the defenders post-snap.

\textbf{den folgenden Teil könnte man auch in den Appendix packen}

A hidden Markov models consists of an observed time series $\{\boldsymbol{y}_t\}_{t=1}^T$ --- here, the y-coordinates of the defensive players and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$ which proxies the offensives players to be guarded at every time point $t$. The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}=\bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij}),$ with $\gamma_{ij} = \Pr(g_t = j| g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $\boldsymbol{y}_t$ are fully determined by the currently active state, i.e.\ 
\begin{equation*}
f(\boldsymbol{y}_t|g_1, \ldots, g_T, \boldsymbol{y}_1, \ldots, \boldsymbol{y}_{t-1},\boldsymbol{y}_{t+1},\ldots,\boldsymbol{y}_T) = f(\boldsymbol{y}_t|g_t).
\end{equation*}
In general, $f$ can be any density or probability mass function depending on the type of data. We are following the approaches of Franks et al. (2015) and Groom (2024) and opt for the Normal distribution.

To remediate this, we derive the decision of man- or zone coverage from the number of switches for individual players. In particular, a low number of switches when offensive players are in motion indicates man coverage whereas a higher number indicates zone coverage.

@@

## Results

Hier die Animation rein mit den Verbindungen von den decodierten States

## Discussion

@@

## Appendix
